\protect Connecting every bit of knowledge: \\ The structure of Wikipedia's Flrst Link Network

### 2016-04-12 writeup/wikipedia-network.tex
edits up to page 3

\title{\protect
Connecting every bit of knowledge: \\
The structure of Wikipedia's first link network
}

@@

\begin{abstract}
  \protect
Apples, porcupines, and the most obscure Bob Dylan song---is every topic a few clicks from Philosophy? 
Within Wikipedia, the surprising answer is yes: nearly all 
paths lead to Philosophy.
Wikipedia is the largest, most meticulously indexed collection of human knowledge ever amassed. 
More than information about a topic, Wikipedia is a web of naturally emerging relationships.  
By following the first link in each article, we algorithmically construct a directed network of 
all 4.7 million articles: Wikipedia's First Link Network.
Here, we study the English edition of Wikipedia's First Link Network for insight into how the many 
inventions, places, people, objects, and events are related and organized.  
We traverse every path, measuring the accumulation of first links, path lengths, basins, cycles, 
and the influence each article exerts in shaping the network.
We discover scale-free distributions describe path length, 
accumulation, and influence. Far from dispersed, first links disproportionately accumulate 
at a few articles---flowing from specific to general and culminating around fundamental notions such as
Community, State, and Science. 
Philosophy shapes more paths than any other article by two orders of magnitude. 
Curiously, we also observe a gravitation towards topical articles such as 
Health Care and Fossil Fuel. 
These findings enrich our view of the connections and structure of
Wikipedia's ever growing store of knowledge.

@@

@@

Wikipedia is a towering achievement of the modern era. 
At no point in history has a larger or more meticulously indexed collection of human knowledge 
existed.
Wikipedia contains 37 million articles in 283 languages, 
with coverage spanning everything from little known ancient battles to the latest pharmaceutical drugs 
\cite{drugs, stats}.
Demonstrating its relevance to modern inquiry,
Wikipedia is the sixth most visited
site in the world, surpassing 18 billion page views and 26 million edits in a single month
\cite{wiki_edits, wiki_views}.

@@

While these many studies have dissected and fruitfully applied Wikipedia's content,
few have examined the connections among the many articles
\cite{links_that_speak}.
A hyperlink from one Wikipedia article to another
naturally indicates a relationship between the two articles
\cite{relevance}.
The notion that hyperlinks convey information about the content of a
page has proved enormously successful in multiple domains from search engine algorithms 
such as PageRank 
\cite{pagerank} 
to topic classification
\cite{classifier}.
We treat a hyperlink as a mechanism connecting two topics.

The authors of a Wikipedia article choose where and whether to include a 
reference to another Wikipedia article
in the HTML markup.
For example, the authors of the 
\href{https://en.wikipedia.org/wiki/Train}{``Train''}
article
chose
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
among others as relevant articles to reference in describing ``Train''
\cite{wiki_train}
.

@@

By focusing our attention on the main body of an article---excluding
elements in the side bars and headings---we 
systematically capture the core description of a topic.
Within the body text, the first link marks the earliest moment in a topic's introduction where the 
authors choose to directly reference another article. 
While many links reference relevant details, 
the first link is an association within a topic's initial 
description. While 
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
are links detailing particulars, the first link, ``rail transport,''
is the topic the authors associate with ``Train'' in the introduction. 
``Banana'' has a first link to ``fruit,'' ``Bob Dylan'' has a first link to 
``Blowing in the Wind,'' and ``Physics'' has a first link to ``natural science.''
Collectively, the first links provide a pragmatic and interpretable 
means to connect each article to another.

By following the first hyperlink to another article in the English edition of
Wikipedia as of November, 2014, we connect the topic of one article to that of another, ultimately forming a directed network: 
{\it Wikipedia's First Link Network} (FLN).
For methodological details, see section~\ref{Appendix}.

The FLN is a wealth of relations among inventions, places,
figures, objects, and events across space and time.
``Train'' for example, links to a parent node, ``Rail Transport,''
while many child nodes such as ``Steel'' and ``Horsepower'' link to ``Train.''
Unlike previous taxonomies created by individuals
\cite{locke, descartes, aristotle, hist_thesaurus},
the relations in the FLN emerge without a centralized effort 
as the aggregate of each article's authors choice of first link.

Our goal is to study the structure of the FLN for insight into how the information on Wikipedia is organized and related.
We develop metrics to capture 
the dominant features of the FLN's structure.
We measure dynamics of the FLN as a flow, quantifying 
the accumulation of first links around articles 
and the influence an article exerts in shaping the FLN.
Together with cycles, in-degree, depth, and the content of the articles, 
we build our analysis of the relations among the ideas in Wikipedia.

@@

An essential feature of a directed network's structure is the degree distribution 
\cite{newman}. 
The degree distribution has been used to study many phenomena from disease outbreak 
\cite{disease} 
to the dynamics of social networks 
\cite{social_nets}.
The in-degree distribution in the FLN describes how many first links point to a 
particular article. 
Articles with zero in-degree have no references---they are outer leaves in the FLN. 
The in-degree 
provides a way to rank the articles in Wikipedia by the number of first link references.

Inspired by the claim that the majority of first links lead to
``Philosophy''---popularized by an 
\href{https://xkcd.com/903/}{xkcd}
comic and subsequently
discussed in blog posts 
\cite{xkcd, mat_blog, Ilmari_first_links}
---
we holistically study how the FLN  yields a flow of connections.
For example ``Train'' has a first link to 
``Rail Transport,''
 which is itself
an article with a first link to 
``Conveyance of Passengers and Goods,''
and so on.
As shown in Fig.~\ref{fig:Train First Links}, the path starting at ``Train'' 
contains ``Goods,'' ``Economics,'' ``Social Science,''
leading ultimately to ``Philosophy''.

@@

Starting at each article, we construct a path through the FLN, 
to map the flow of connections among articles.
The method is order agnostic with respect to which articles are selected first. As long as each article is selected eventually, the resulting metrics are equivalent.
Previous studies have used flow to characterize the structure of river networks
\cite{geo_basins, dodds} and describe the organization of food systems through transportation networks
\cite{food_webs}.
With paths to mark flow through the FLN, we develop metrics to 
gauge the accumulation of references, 
the length of the path relating articles, and the influence a particular
article exerts in shaping the flow of links through the FLN.

The first metric we develop quantifies the accumulation of first links.
The algorithm begins by selecting an article, then traversing the path formed
by following the first links. Each time a first link references an article, we increment a count
associated with the article. 
We continue until the first link is retraced or is invalid---defined here to mean outside of Wikipedia.
We select a second article and repeat the process until we have 
constructed a path for each article in the network. We call the resulting count for each article the number of {\it traversal visits}. The number of traversal visits of an article 
measures the number of references flowing to the ideas in the article---equivalent
to basin area in geomorphology.

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/traversal_funnel_algo_figure.pdf}
  \caption{
    \textbf{Traversal Funnel Algorithm on a sample network.}
  The algorithm for traversal funnels is identical to the previous algorithm for traversal visits with one alteration: the path ends at the start of a cycle to distinguish articles directing a path into a cycle from articles that simply happen to be in a highly traversed path. We can construct similar vectors by considering each path through the network, measuring traversal funnels for a particular article as the sum of the entries in its corresponding row. For example
  the number of traversal funnels for article $E$ is 
  $\sum_{i=1}^7 E_{\text{funnel, i}} = 2$.}
  \label{fig:Traversal Funnels}

@@

>>>>

\title{\protect
Connecting every bit of knowledge: \\
The structure of Wikipedia's Flrst Link Network
}

@@

\begin{abstract}
  \protect
Apples, porcupines, and the most obscure Bob Dylan song---is every topic a few clicks from Philosophy? 
Within Wikipedia, the surprising answer is yes: nearly all 
paths lead to Philosophy.
Wikipedia is the largest, most meticulously indexed collection of human knowledge ever amassed. 
More than information about a topic, Wikipedia is a web of naturally emerging relationships.  
By following the first link in each article, we algorithmically construct a directed network of 
all 4.7 million articles: Wikipedia's First Link Network.
Here, we study the English edition of Wikipedia's First Link Network for insight into how the many 
articles on inventions, places, people, objects, and events are related and organized.  
We traverse every path, measuring the accumulation of first links, path lengths, basins, cycles, 
and the influence each article exerts in shaping the network.
We find scale-free distributions describe path length, 
accumulation, and influence. Far from dispersed, first links disproportionately accumulate 
at a few articles---flowing from specific to general and culminating around fundamental notions such as
Community, State, and Science. 
Philosophy funnels more paths than any other article by two orders of magnitude. 
We also observe a gravitation towards topical articles such as 
Health Care and Fossil Fuel. 
These findings enrich our view of the connections and structure of
Wikipedia's ever growing store of knowledge.

@@

%TODO:
% Abstract 
    % In traversing every path, we measure ?
    % Our findings... braided network (what does this mean?)
% Formatting
    % incorporate into template
% Intro
    % add month to monthly view stats
% Method
    % what to cite for drainage basin in geomorphology

% continue on Page 3

@@

Wikipedia is a towering achievement of the Internet age. 
At no point in history has a larger or more meticulously indexed collection of human knowledge 
existed.
Wikipedia contains 37 million articles in 283 languages, 
with coverage spanning everything from little known ancient battles to the latest pharmaceutical drugs 
\cite{drugs, stats}.
Demonstrating its relevance to modern inquiry,
Wikipedia is the sixth most visited
site in the world, surpassing 18 billion page views and 11 million edits in a single month.
\cite{wiki_edits, wiki_views}.

@@

While these many studies have dissected and fruitfully applied Wikipedia's content,
here we examine the connections among the many articles.
A hyperlink from one Wikipedia article to another
naturally indicates a relationship between the two articles
\cite{relevance}.
The notion that hyperlinks convey information about the content of a
page has proved enormously successful in multiple domains from search engine algorithms 
such as PageRank 
\cite{pagerank} 
to topic classification
\cite{classifier}.
Here, we treat a hyperlink as a mechanism connecting two topics.

The authors of a Wikipedia article choose where and whether to include a 
reference to another Wikipedia article
in the HTML markup.
For example, as of November 2014, the authors of the 
\href{https://en.wikipedia.org/wiki/Train}{``Train''}
article
had collectively chosen
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
among others as relevant articles to reference in describing ``Train''
\cite{wiki_train}
.

@@

By focusing our attention on the main body of an article---excluding
elements in side bars and headings---we 
attempt to systematically capture the core description of a topic.
Within the body text, the first link marks the earliest moment in a topic's introduction where the 
authors choose to directly reference another article. 
While many links reference relevant details, 
the first link is an association within a topic's initial 
description. While 
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
are links detailing particulars, the first link, ``rail transport,''
is the topic the authors associate with ``Train'' in the introduction
(Fig.~\ref{fig:Train First Links}).
``Banana'' has a first link to ``fruit,'' ``Bob Dylan'' has a first link to 
``Blowing in the Wind,'' and ``Physics'' has a first link to ``natural science.''
Collectively, first links provide a pragmatic and interpretable 
means to connect each article to another.

By following the first hyperlink contained in all articles of the English edition of
Wikipedia as of November, 2014, we form a directed network:
{\it Wikipedia's First Link Network} (FLN).
Inspired by the claim that the majority of first links lead to
``Philosophy''---popularized by an 
\href{https://xkcd.com/903/}{xkcd}
comic and subsequently
discussed in blog posts 
\cite{xkcd, mat_blog, Ilmari_first_links}
---
we holistically study how the FLN  yields a flow of connections.
For methodological details, see Sec.~\ref{Appendix}.

The FLN is a wealth of relations among inventions, places,
figures, objects, and events across space and time.
``Train'' for example, links to a parent node, ``Rail Transport,''
while many child nodes such as ``Steel'' and ``Horsepower'' link to ``Train.''
As shown in Fig.~\ref{fig:Train First Links}, the path starting at ``Train'' 
contains ``Goods,'' ``Economics,'' ``Social Science,''
leading ultimately to ``Philosophy''.
Unlike previous taxonomies created by individuals
\cite{locke, descartes, aristotle, hist_thesaurus},
the relations in the FLN emerge without a centralized effort 
as the aggregate of each article's authors choice of first link.

Our goal is to study the structure of the FLN for insight into how the information on Wikipedia is organized and related.
Informed by river network metrics,
we consider the FLN as a flow, quantifying 
the accumulation of first links around articles 
and the influence an article exerts in shaping the FLN.
Together with cycles, in-degree, depth, and the content of the articles, 
we build our analysis of the relations among the ideas in Wikipedia.

@@

An essential feature of a directed network's structure is the degree distribution 
\cite{newman}. 
The degree distribution has been used to study many phenomena from disease outbreak 
\cite{disease} 
to the dynamics of social networks 
\cite{social_nets}.
The in-degree distribution in the FLN describes how many first links point to a 
particular article. 
Articles with zero in-degree have no references---they are outer leaves in the FLN 
or are disconnected entirely.

@@

Starting at each article, we construct a path through the FLN, 
to map the flow of connections among articles.
The method is order agnostic with respect to which articles are selected first. As long as each article is selected eventually, the resulting metrics are equivalent.
Previous studies have used flow to characterize the structure of river networks
\cite{geo_basins, dodds} and describe the organization of 
blood networks, food systems, and transportation networks
\cite{food_webs}.

The first metric we develop quantifies the accumulation of first links.
The algorithm begins by selecting an article, then traversing the path formed
by following the first links. Each time a first link references an article, we increment a count
associated with the article. 
We continue until a page is revisited or is invalid---defined here to mean pointing to a page outside of Wikipedia.
We select a second article and repeat the process until we have 
constructed a path for each article in the network. 
We define the number of {\it traversal visits} of an article 
to be the number of references flowing to the ideas in the article---equivalent
to drainage basin area in geomorphology.

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/traversal_funnel_algo_figure.pdf}
  \caption{
    \textbf{Traversal Funnel Algorithm on a simple network.}
  The algorithm for traversal funnels is identical to the previous algorithm for traversal visits with one alteration: the path ends at the start of a cycle to distinguish articles directing a path into a cycle from articles that simply happen to be in a highly traversed path. We can construct similar vectors by considering each path through the network, measuring traversal funnels for a particular article as the sum of the entries in its corresponding row. For example
  the number of traversal funnels for article $E$ is 
  $\sum_{i=1}^7 E_{\text{funnel, i}} = 2$.}
  \label{fig:Traversal Funnels}

@@

The authors are also grateful for the suggestions and input provided by Randall Harp.
